---
title: "COVID-19: The Daily Report"
date: "6/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(scales)

# The NY Times shares information on the total number of cases in the US:
url_total = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us.csv"
covid_data <- readr::read_csv(url_total)

# as well as split by state:
url_by_state = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
covid_data_by_state <- readr::read_csv(url_by_state)
```

This Rmarkdown file is intended to give a visual snapshot of the progression of coronavirus: by state, by region, or in total. Data is graciously provided by the NY Times [(accessible via Github)](https://github.com/nytimes/covid-19-data). Source code for this file can be found on [Github](https://github.com/brianrice2/covid-19).

The most recent date of available information is `r format(max(covid_data$date), "%B %d, %Y")`.

### National overview - new cases and deaths

```{r new-daily-cases}
# new daily cases
covid_data$new_cases <- c(0, diff(covid_data$cases))
ggplot(data = covid_data) +
  geom_line(mapping = aes(x = date, y = new_cases),
            color = 'steelblue') +
  labs(x = 'Date', y = 'New daily cases', title = 'Daily cases') +
  scale_y_continuous(labels = comma)
```


```{r new-daily-deaths}
# new daily deaths
covid_data$new_deaths <- c(0, diff(covid_data$deaths))
ggplot(data = covid_data) +
  geom_line(mapping = aes(x = date, y = new_deaths),
            color = 'red4') +
  labs(x = 'Date', y = 'New daily deaths', title = 'Daily deaths') +
  scale_y_continuous(labels = comma)
```

### State-level snapshots

```{r state-setup}
# add new daily cases and deaths to state data
covid_data_by_state <- covid_data_by_state %>% 
  group_by(state) %>% 
  arrange(date) %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

covid_data_by_state <- covid_data_by_state %>% 
  replace_na(list(new_cases = 0, new_deaths = 0))
```

```{r all-states-new-cases, fig.width=10, fig.height=24}
# view new cases for all states (still split by state)
covid_data_by_state %>% 
  ggplot() +
    geom_line(mapping = aes(x = date, y = new_cases, color = state)) +
    facet_wrap(~ state, nrow = 10, shrink = FALSE) +
    labs(x = 'Date', y = 'New daily cases') +
    scale_y_continuous(labels = comma) +
    theme(legend.position = "none")
```

### Regional progression

The US Census Bureau provides regional classifications to us - [link](https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf) - which I will use. It's 1) easy, and 2) seems good enough to identify general trends. Notably, the "other" category includes Hawaii, Alaska, Puerto Rico, and the Northern Mariana Islands.

```{r region-classification}
west <- c('Washington', 'Oregon', 'California', 'Nevada', 'Idaho',
          'Utah', 'Arizona', 'Montana', 'Wyoming', 'Colorado', 'New Mexico')
midwest <- c('North Dakota', 'South Dakota', 'Nebraska', 'Kansas', 'Minnesota',
             'Iowa', 'Missouri', 'Wisconsin', 'Illinois', 'Indiana',
             'Michigan', 'Ohio')
south <- c('Oklahoma', 'Texas', 'Arkansas', 'Louisiana', 'Mississippi',
           'Tennessee', 'Kentucky', 'Alabama', 'Georgia', 'West Virginia',
           'Virginia', 'North Carolina', 'South Carolina', 'Florida',
           'Maryland', 'Delaware', 'District of Columbia')
northeast <- c('Pennsylvania', 'New Jersey', 'New York', 'Vermont', 'New Hampshire',
               'Massachusetts', 'Connecticut', 'Rhode Island', 'Maine')
```

```{r assign-groups}
covid_data_by_state$region <- with(covid_data_by_state,
                                   ifelse(state %in% west, "west",
                                          ifelse(state %in% midwest, "midwest",
                                                 ifelse(state %in% south, "south",
                                                        ifelse(state %in% northeast, "northeast", "other")))))
```

```{r plot-new-regional-cases, message=FALSE}
covid_data_by_state %>% 
  group_by(region, date) %>%
  summarize(regional_new_cases = sum(new_cases)) %>% 
  ggplot() +
    geom_line(mapping = aes(x = date, y = regional_new_cases), color = 'steelblue') +
    facet_wrap(~ region, nrow = 2) +
    labs(x = 'Date', y = 'New daily cases', title = 'Daily cases') +
    scale_y_continuous(labels = comma)
```

```{r plot-new-regional-deaths, message=FALSE}
covid_data_by_state %>% 
  group_by(region, date) %>%
  summarize(regional_new_deaths = sum(new_deaths)) %>% 
  ggplot() +
    geom_line(mapping = aes(x = date, y = regional_new_deaths), color = 'red4') +
    facet_wrap(~ region, nrow = 2) +
    labs(x = 'Date', y = 'New daily deaths', title = 'Daily deaths') +
    scale_y_continuous(labels = comma)
```

### Just the past two weeks

```{r plot-new-regional-cases-past-two-weeks, message=FALSE}
covid_data_by_state %>%
  filter(date >= max(date) - 14) %>%
  group_by(region, date) %>%
  summarize(regional_new_cases = sum(new_cases)) %>%
  ggplot() +
    geom_line(mapping = aes(x = date, y = regional_new_cases), color = 'steelblue') +
    facet_wrap(~ region, nrow = 2) +
    labs(x = 'Date', y = 'New daily cases', title = 'Daily cases') +
    scale_y_continuous(labels = comma)
```

```{r plot-new-regional-deaths-past-two-weeks, message=FALSE}
covid_data_by_state %>%
  filter(date >= max(date) - 14) %>%
  group_by(region, date) %>%
  summarize(regional_new_deaths = sum(new_deaths)) %>%
  ggplot() +
    geom_line(mapping = aes(x = date, y = regional_new_deaths), color = 'red4') +
    facet_wrap(~ region, nrow = 2) +
    labs(x = 'Date', y = 'New daily deaths', title = 'Daily deaths') +
    scale_y_continuous(labels = comma)
```

### Cases per capita

The US Census Bureau provides [estimates of each state's population as of 2019](https://www2.census.gov/programs-surveys/popest/datasets/). Let's pull that info in so we can look at new cases per capita

```{r population-data-setup, message=FALSE}
population_data <- read_csv('./census-population-estimates.csv')
population_data <- population_data %>% 
  select(NAME, POPESTIMATE2019)

population_data <- population_data %>%
  rename(state = NAME,
         population = POPESTIMATE2019)

# add population data to our original covid data
covid_data_by_state <- left_join(covid_data_by_state, population_data, by = 'state')

# calculate cases and deaths per capita
covid_data_by_state <- mutate(covid_data_by_state,
                              new_cases_per_capita = new_cases / population,
                              new_deaths_per_capita = new_deaths / population)
```

```{r plot-new-regional-cases-per-capita-past-two-weeks, message=FALSE}
covid_data_by_state %>%
  filter(date >= max(date) - 14) %>%
  group_by(region, date) %>%
  summarize(regional_new_cases_per_capita = sum(new_cases_per_capita)) %>%
  ggplot() +
    geom_line(mapping = aes(x = date, y = regional_new_cases_per_capita), color = 'steelblue') +
    facet_wrap(~ region, nrow = 2) +
    labs(x = 'Date', y = 'New daily cases', title = 'Daily cases as a percentage of population') +
    scale_y_continuous(labels = percent)
```

```{r plot-new-regional-deaths-per-capita-past-two-weeks, message=FALSE}
covid_data_by_state %>%
  filter(date >= max(date) - 14) %>%
  group_by(region, date) %>%
  summarize(regional_new_deaths_per_capita = sum(new_deaths_per_capita)) %>%
  ggplot() +
    geom_line(mapping = aes(x = date, y = regional_new_deaths_per_capita), color = 'red4') +
    facet_wrap(~ region, nrow = 2) +
    labs(x = 'Date', y = 'New daily deaths', title = 'Daily deaths as a percentage of population') +
    scale_y_continuous(labels = percent)
```
